---
title: "Final Project: Regression and Classification with Multi Architecture (Deep) Neural Networks"
author: "EMMANUEL OPOKU(22630361)"
format: pdf
execute:
  echo: false
  warning: false
  message: false
---

# REPRODUCIBILITY

```{r}
# =========================
# REPRODUCIBILITY
# =========================
set.seed(260408)

```

# LIBRARIES

```{r}
# =========================
# LIBRARIES
# =========================
library(tidyverse)
library(lubridate)
library(corrplot)
library(ggplot2)
library(forecast)
library(tseries)
library(keras3)
library(tensorflow)
```

# **Introduction**

The current research analyses models based on the following types of neural networks: deep neural networks (DNNs), convolutional neural networks (CNNs), long short-term memory networks (LSTMs), and hybrid networks. For regression analysis, the study utilises the Jena Climate dataset, while for the image classification task, the MNIST database will be used. The overall goal of this investigation is to analyse the effect of certain architectural features of neural networks on their ability to learn.

# Part 0: Data Preparation and Exploratory Analysis

```{r}
jena <- read.csv(choose.files(), h=T)
# Check structure
str(jena)
# Preview
head(jena)
# Row count BEFORE resampling
rows_before <- nrow(jena)

# PARSE DATE-TIME
# =========================
jena$Date.Time <- dmy_hms(jena$Date.Time)

# RESAMPLE TO HOURLY DATA
# =========================
jena_hourly <- jena %>%
  mutate(Date.Hour = floor_date(Date.Time, "hour")) %>%
  group_by(Date.Hour) %>%
  summarise(
    across(where(is.numeric), mean, na.rm = TRUE),
    .groups = "drop"
  )

# Row count AFTER resampling
rows_after <- nrow(jena_hourly)
# Display results
rows_before
rows_after
head(jena_hourly)
str(jena_hourly)
```

# **The Jena Climate dataset**

The Jena Climate dataset was loaded from the provided CSV file, while the DateTime column was transformed into a proper date-time variable with the help of the lubridate package. The dataset contains data points taken every 10 minutes.

In order to meet the requirements for model development, the data had to be resampled with an hourly frequency through aggregation, meaning the average of all numeric variables was calculated for each hour. It is important to note that the resampling process produces one observation per hour, which makes it possible to use a 120-hour lookback window when building neural networks.

Before the data transformation took place, there were r rows_before observations in the data. After the resampling process, there were only r rows_after observations left.

# CHECK MISSING VALUES

```{r}
# =========================
# CHECK MISSING VALUES
# =========================
missing_values <- colSums(is.na(jena_hourly))
missing_values

```

# **Missing Values Analysis**

For the presence of missing values in the data set, the total number of missing values in each variable was counted.

The results show that after going through the resampling process with a frequency of one hour, there are no major missing values among the variables.

Since the resampling stage included mean aggregation with na.rm=TRUE, the presence of minor missing values within the data set before the resampling process would have been addressed by aggregation. Thus, imputation was not necessary in this situation.

This is true since imputations such as mean imputation would bring about biases while aggregation maintains the validity of the data, ensuring that the overall trends and patterns in the dataset are preserved without introducing artificial values.

# FULL TIME SERIES PLOT

```{r}
# =========================
# FULL TIME SERIES PLOT
# =========================
ggplot(jena_hourly, aes(x = Date.Hour, y = T..degC.)) +
  geom_line(color = "green") +
  labs(title = "Full Temperature Time Series",
       x = "Time",
       y = "Temperature (°C)") +
  theme_minimal()

```

# FIRST 10 DAYS PLOT

```{r}
# =========================
# FIRST 10 DAYS PLOT
# =========================
first_10_days <- jena_hourly %>%
  filter(Date.Hour <= min(Date.Hour) + days(10))

ggplot(first_10_days, aes(x = Date.Hour, y = T..degC.)) +
  geom_line(color = "blue") +
  labs(title = "First 10 Days Temperature (Hourly)",
       x = "Time",
       y = "Temperature (°C)") +
  theme_minimal()
  
```

# **Time Series Visualization**

There is an evident presence of long-term seasonal patterns in the whole series of temperature, which is evident from the oscillatory nature of temperature with each successive year. These are due to annual climatic trends where temperatures will be higher in warm months and lower in cold months.

A review of the first 10 days' hourly temperature series suggests the presence of short-term variations, with a notable daily pattern where temperatures increase during the day and decrease during nighttime.

In view of the above, the need to consider a 120-hour lookback period makes sense because it captures short-term as well as long-term temporal dependencies in the data set.

# CORRELATION MATRIX

```{r}
# =========================
# CORRELATION MATRIX
# =========================
numeric_data <- jena_hourly %>%
  select(where(is.numeric))

cor_matrix <- cor(numeric_data)

# Plot correlation matrix
corrplot(cor_matrix, method = "color", type = "upper",
         tl.cex = 0.7, tl.col = "black")
```

# Correlation Analysis and Multicollinearity

The correlation matrix shows the connections between all predictors used in the data set. Some of the predictors have shown to exhibit strong positive or negative correlation among themselves, especially temperature predictors like T..degC, Tpot..K, and Tdew..degC, and others are vapour pressure-related predictors.

Such behaviour implies that there is a case of multicollinearity since some of the predictors are providing redundant information. Multicollinearity may pose a problem for statistical models, but neural networks appear to be more tolerant of redundancy because they can identify nonlinear dependencies.

Still, it is essential to know the details because some correlations might cause problems in training the model.

# ACF AND PACF OF TEMPERATURE

```{r}
# =========================
# ACF AND PACF OF TEMPERATURE
# =========================
temperature_series <- jena_hourly$T..degC.

acf(temperature_series,
    lag.max = 168,
    main = "ACF of Hourly Temperature Series")

pacf(temperature_series,
     lag.max = 168,
     main = "PACF of Hourly Temperature Series")

```

# **ACF and PACF Analysis**

Based on the autocorrelation function (ACF), there are significant levels of correlation between temperature values and prior hours over an extended period of time. The slow decline in autocorrelation indicates that there is substantial serial correlation in the data set. This is expected since the current value of temperature is dependent on its previous values.

The partial autocorrelation function (PACF) shows a large number of partial correlations up to certain lags, which shows that the previous hours contain valuable information about prediction. Therefore, a lookback window of 120 hours should be adopted to use the past 5 days of weather conditions to predict temperature after 24 hours.

# **BOXPLOTS OF PREDICTORS**

```{r}
# =========================
# BOXPLOTS OF PREDICTORS
# =========================
predictor_data <- jena_hourly %>%
  select(-Date.Hour)

predictor_long <- predictor_data %>%
  pivot_longer(cols = everything(),
               names_to = "Variable",
               values_to = "Value")

ggplot(predictor_long, aes(x = Variable, y = Value)) +
  geom_boxplot() +
  coord_flip() +
  labs(title = "Boxplots of Jena Climate Predictor Variables",
       x = "Predictor Variables",
       y = "Observed Values") +
  theme_minimal()
```

# **Predictor Range and Need for Normalization**

From the box plots above, it is evident that the features have been captured on very different scales. This is because atmospheric pressure, temperature, humidity, wind direction, and vapour pressure have different units of measure, as well as different numerical values.

This confirms the need for data normalisation before training any neural network models. Neural networks tend to be sensitive to differences in feature scaling because of how gradient descent functions. It will work inefficiently if there are very different scales for the features.

# CHRONOLOGICAL TRAIN/VALIDATION/TEST SPLIT

```{r}
# =========================
# CHRONOLOGICAL TRAIN/VALIDATION/TEST SPLIT
# =========================
train_data <- jena_hourly %>%
  filter(Date.Hour >= ymd_hms("2009-01-01 00:00:00"),
         Date.Hour <  ymd_hms("2015-01-01 00:00:00"))

val_data <- jena_hourly %>%
  filter(Date.Hour >= ymd_hms("2015-01-01 00:00:00"),
         Date.Hour <  ymd_hms("2016-01-01 00:00:00"))

test_data <- jena_hourly %>%
  filter(Date.Hour >= ymd_hms("2016-01-01 00:00:00"),
         Date.Hour <  ymd_hms("2017-01-01 00:00:00"))

# Select numeric columns only
numeric_cols <- names(jena_hourly %>% select(where(is.numeric)))

# Compute training-set mean and standard deviation only
train_mean <- train_data %>%
  select(all_of(numeric_cols)) %>%
  summarise(across(everything(), mean, na.rm = TRUE))

train_sd <- train_data %>%
  select(all_of(numeric_cols)) %>%
  summarise(across(everything(), sd, na.rm = TRUE))

# Function for z-score normalization
normalize_with_train_stats <- function(data, mean_values, sd_values) {
  data_scaled <- data
  
  data_scaled[numeric_cols] <- sweep(
    data_scaled[numeric_cols],
    2,
    as.numeric(mean_values),
    FUN = "-"
  )
  
  data_scaled[numeric_cols] <- sweep(
    data_scaled[numeric_cols],
    2,
    as.numeric(sd_values),
    FUN = "/"
  )
  
  return(data_scaled)
}

# Apply normalization
train_scaled <- normalize_with_train_stats(train_data, train_mean, train_sd)
val_scaled   <- normalize_with_train_stats(val_data, train_mean, train_sd)
test_scaled  <- normalize_with_train_stats(test_data, train_mean, train_sd)

# Check normalized training data
summary(train_scaled %>% select(where(is.numeric)))
```

# **Z-score Normalization**

Z-score normalisation was conducted for all numerical features, employing only the means and standard deviations obtained from the training dataset. Validation and test datasets were then normalised based on the aforementioned statistics.

It is crucial to do so since validation and test datasets are supposed to resemble unknown future data. Utilisation of their means and standard deviations in normalisation will allow data from the future to seep into the training process. It could result in overly optimistic validation and test performance since the algorithm will be indirectly trained on future data.

# CREATE 24-HOUR-AHEAD TEMPERATURE TARGET

```{r}
# =========================
# CREATE 24-HOUR-AHEAD TEMPERATURE TARGET
# =========================
jena_hourly <- jena_hourly %>%
  arrange(Date.Hour) %>%
  mutate(
    temp_24h_ahead = lead(T..degC., 24),
    temp_class = case_when(
      temp_24h_ahead < 0 ~ "Cold",
      temp_24h_ahead >= 0  & temp_24h_ahead < 10 ~ "Cool",
      temp_24h_ahead >= 10 & temp_24h_ahead < 20 ~ "Mild",
      temp_24h_ahead >= 20 ~ "Warm",
      TRUE ~ NA_character_
    )
  )

# Remove rows with missing future target
jena_class <- jena_hourly %>%
  filter(!is.na(temp_24h_ahead),
         !is.na(temp_class))

# Class frequencies
class_freq <- jena_class %>%
  count(temp_class) %>%
  mutate(percentage = round(100 * n / sum(n), 2))

class_freq
```

# BAR PLOT OF CLASS FREQUENCIES

```{r}
# =========================
# BAR PLOT OF CLASS FREQUENCIES
# =========================
ggplot(class_freq, aes(x = temp_class, y = n)) +
  geom_col() +
  labs(title = "Class Frequencies for 24-Hour-Ahead Temperature",
       x = "Temperature Class",
       y = "Frequency") +
  theme_minimal()
```

# **Derived Temperature Classification Label**

Four classes were used in the construction of the classification target based on the 24-hour-ahead temperature, which included 'Cold' for temperatures less than 0°C, 'Cool' for temperatures between 0°C and 9.9°C, 'Mild' for temperatures between 10°C and 19.9°C, and 'Warm' for temperatures of 20°C and above.

The class frequency chart shows how each of the four temperature classes is distributed. If some of the classes have more instances compared to others, then the task is said to be class-imbalanced. The reason why this topic is important is that the classification model might achieve high accuracy in predicting the most frequent class but still perform poorly in predicting other classes.

# DATA GENERATOR FUNCTION

```{r}
# =========================
# DATA GENERATOR FUNCTION
# =========================

data_generator <- function(data,
                           lookback = 120,
                           delay = 24,
                           step = 1,
                           batch_size = 128,
                           shuffle = FALSE,
                           task = "regression") {
  
  # Extract numeric predictors
  data_values <- as.matrix(data %>% select(where(is.numeric)))
  
  max_index <- nrow(data_values) - delay - 1
  
  function() {
    
    if (shuffle) {
      rows <- sample((lookback + 1):max_index, size = batch_size)
    } else {
      rows <- seq(lookback + 1, min(lookback + batch_size, max_index))
    }
    
    samples <- array(0, dim = c(length(rows), lookback, ncol(data_values)))
    
    if (task == "classification") {
      targets <- matrix(0, nrow = length(rows), ncol = 4)
    } else {
      targets <- numeric(length(rows))
    }
    
    for (j in seq_along(rows)) {
      idx <- rows[j]
      indices <- seq(idx - lookback, idx - 1, by = step)
      
      samples[j,,] <- data_values[indices,]
      
      if (task == "regression") {
        targets[j] <- data$T..degC.[idx + delay]
      } else {
        class_label <- data$temp_class[idx + delay]
        targets[j,] <- as.numeric(factor(class_label,
                                         levels = c("Cold","Cool","Mild","Warm")))
      }
    }
    
    list(samples, targets)
  }
}
```

# GENERATORS FOR EACH SPLIT

```{r}
# =========================
# GENERATORS FOR EACH SPLIT
# =========================

train_gen <- data_generator(train_scaled, shuffle = TRUE, task = "regression")
val_gen   <- data_generator(val_scaled, shuffle = FALSE, task = "regression")
test_gen  <- data_generator(test_scaled, shuffle = FALSE, task = "regression")

```

# VERIFY SHAPES

```{r}
# =========================
# VERIFY SHAPES
# =========================
batch <- train_gen()

dim(batch[[1]])  # samples
dim(batch[[2]])  # targets
```

# **Expected Output**

samples: (batch_size, 120, number_of_features) targets: (batch_size) → regression targets: (batch_size, 4) → classification

# **Data Generator Function**

The custom data generator function was implemented to generate input and output sequences suitable for use in neural network training. This function receives arguments such as look back (120 hours), delay (24 hours), step size, batch size, shuffling, and task type (regression or classification).

For each data sample, the generator uses a sliding window approach where it fetches past data samples within a span of 120 hours. These input samples are then paired with the target value that denotes the temperature value after 24 hours.

In case of regression problems, the target would be the actual numeric value of temperature. In classification problems, the target will be the class label that represents the temperature category (cold, cool, mild, or warm).

As can be observed, the generated input data will be three-dimensional in nature, i.e., (batch size, time steps, number of features). This data structure will be useful while training deep learning models such as DNN, CNN, and LSTM networks. As evident from the output shapes, the custom data generator produces correct input data.

It is worth mentioning that the memory-efficient and scalable implementation strategy was used, where data was generated using batches rather than loading complete data in memory.

# LOAD MNIST DATASET

```{r}
# =========================
# LOAD MNIST DATASET
# =========================
mnist <- dataset_mnist()

x_train_raw <- mnist$train$x
y_train_raw <- mnist$train$y

x_test_raw <- mnist$test$x
y_test_raw <- mnist$test$y

# Verify dimensions
dim(x_train_raw)
dim(x_test_raw)

# Check pixel value range
range(x_train_raw)
range(x_test_raw)
```

# **MNIST Data Loading**

The MNIST handwritten digit dataset was loaded using dataset_mnist() `from Keras`. The training set contains 60,000 greyscale images, while the test set contains 10,000 images. Each image has dimensions of 28 × 28 pixels. The pixel values range from 0 to 255 before normalisation.

# NORMALIZE PIXELS TO \[0, 1\]

```{r}
# =========================
# NORMALIZE PIXELS TO [0, 1]
# =========================
x_train <- x_train_raw / 255
x_test  <- x_test_raw / 255

# =========================
# VALIDATION SPLIT
# Last 10,000 training images used for validation
# =========================
x_val <- x_train[50001:60000,,]
y_val <- y_train_raw[50001:60000]

x_train_final <- x_train[1:50000,,]
y_train_final <- y_train_raw[1:50000]

# Verify split dimensions
dim(x_train_final)
dim(x_val)
dim(x_test)
```

# DISPLAY SAMPLE MNIST IMAGES

```{r}
# =========================
# DISPLAY SAMPLE MNIST IMAGES
# =========================
par(mfrow = c(5, 2), mar = c(1, 1, 2, 1))

for (i in 1:10) {
  image(
    t(apply(x_train_final[i,,], 2, rev)),
    col = gray.colors(256),
    axes = FALSE,
    main = paste("Label:", y_train_final[i])
  )
}
```

# **Sample MNIST Images**

A 5 × 2 grid of sample MNIST images was displayed together with their corresponding digit labels. The images show handwritten digits from 0 to 9 in greyscale format. This confirms that the image-label pairs were correctly loaded.

# CLASS FREQUENCIES

```{r}
# =========================
# CLASS FREQUENCIES
# =========================
mnist_class_freq <- as.data.frame(table(y_train_final))

colnames(mnist_class_freq) <- c("Digit", "Frequency")

mnist_class_freq$Percentage <- round(
  100 * mnist_class_freq$Frequency / sum(mnist_class_freq$Frequency),
  2
)

mnist_class_freq
```

# BAR PLOT OF CLASS FREQUENCIES

```{r}
# =========================
# BAR PLOT OF CLASS FREQUENCIES
# =========================
ggplot(mnist_class_freq, aes(x = Digit, y = Frequency)) +
  geom_col() +
  labs(title = "MNIST Training Class Frequencies",
       x = "Digit Class",
       y = "Frequency") +
  theme_minimal()
```

# I.B Classification DNN (MNIST)

```{r}
# =========================
# DNN FORMAT: 784-LENGTH VECTOR
# =========================
x_train_dnn <- array_reshape(x_train_final, c(nrow(x_train_final), 28 * 28))
x_val_dnn   <- array_reshape(x_val, c(nrow(x_val), 28 * 28))
x_test_dnn  <- array_reshape(x_test, c(nrow(x_test), 28 * 28))

# =========================
# CNN FORMAT: 28 x 28 x 1
# =========================
x_train_cnn <- array_reshape(x_train_final, c(nrow(x_train_final), 28, 28, 1))
x_val_cnn   <- array_reshape(x_val, c(nrow(x_val), 28, 28, 1))
x_test_cnn  <- array_reshape(x_test, c(nrow(x_test), 28, 28, 1))

# =========================
# ONE-HOT ENCODE LABELS
# =========================
y_train_cat <- to_categorical(y_train_final, num_classes = 10)
y_val_cat   <- to_categorical(y_val, num_classes = 10)
y_test_cat  <- to_categorical(y_test_raw, num_classes = 10)

# Verify shapes
dim(x_train_dnn)
dim(x_train_cnn)
dim(y_train_cat)
```

# **Input Transformations for Neural Networks**

In the DNN baseline model, the images of dimensions `28 × 28` were reshaped to form a vector of size 784. This arrangement is ideal since fully connected layers expect inputs in a one-dimensional form.

In the CNN model, the original spatial arrangement of the images was maintained. This is achieved through rearrangement of each image to produce a tensor of shape `28 × 28 × 1`. The third dimension is responsible for the single greyscale channel. In this case, the convolution layers can extract local patterns like edges and digit curves.

The digit labels were also converted into one-hot encoded vectors with 10 classes, which is required when using categorical cross-entropy loss with a softmax output layer.

# DNN REGRESSION MODEL FOR JENA

```{r}
# =========================
# DNN REGRESSION MODEL FOR JENA
# =========================

# Number of predictors from generator batch
sample_batch <- train_gen()
num_features <- dim(sample_batch[[1]])[3]

# Define DNN model
dnn_reg_model <- keras_model_sequential() %>%
  layer_flatten(input_shape = c(120, num_features)) %>%
  layer_dense(units = 128, activation = "relu") %>%
  layer_dense(units = 64, activation = "relu") %>%
  layer_dense(units = 1, activation = "linear")

# Model summary
summary(dnn_reg_model)
```

# **Regression DNN Architecture**

A vanilla DNN was used as the baseline architecture to solve the Jena temperature forecasting problem. Since the input features are a combination of 120 hours of observation on each of the predictors, the first step involves flattening the inputs into a one-dimensional vector before feeding it into the dense layers.

The two hidden layers have ReLU activation functions. ReLU was chosen due to its ability to introduce non-linearity to the model without the issue of vanishing gradient associated with sigmoid and tanh. The number of units in the first hidden layer is 128, while that of the second hidden layer is 64.

The output layer has only one neurone whose activation is linear. A linear activation makes sense for a regression problem because the temperature target values are continuous and may have negative values. In regression problems, there is no need for an output that represents probabilities; therefore, the linear neurone helps predict unrestricted real-valued target temperatures.

# HELPER FUNCTION: PARAMETER COUNT

```{r}
# =========================
# HELPER FUNCTION: PARAMETER COUNT
# =========================
count_trainable_params <- function(model) {
  sum(sapply(model$trainable_weights, keras::k_count_params))
}

# =========================
# TRAINING SETTINGS
# =========================
batch_size <- 128
steps_per_epoch <- 200
validation_steps <- 50
epochs <- 50

early_stop <- callback_early_stopping(
  monitor = "val_mae",
  patience = 5,
  restore_best_weights = TRUE
)
```

# 1. DNN Without Regularization

```{r}
# =========================
# DNN MODEL 1: NO REGULARIZATION
# =========================
dnn_no_reg <- keras_model_sequential() %>%
  layer_flatten(input_shape = c(120, num_features)) %>%
  layer_dense(units = 128, activation = "relu") %>%
  layer_dense(units = 64, activation = "relu") %>%
  layer_dense(units = 1, activation = "linear")

dnn_no_reg %>% compile(
  optimizer = optimizer_adam(),
  loss = "mae",
  metrics = c("mae")
)

summary(dnn_no_reg)

params_no_reg <- count_trainable_params(dnn_no_reg)
params_no_reg
```

```{r}
history_no_reg <- dnn_no_reg %>% fit(
  train_gen,
  steps_per_epoch = steps_per_epoch,
  epochs = epochs,
  validation_data = val_gen,
  validation_steps = validation_steps,
  callbacks = list(early_stop)
)
```

# 2. DNN With Dropout

```{r}
# =========================
# DNN MODEL 2: DROPOUT
# =========================
dnn_dropout <- keras_model_sequential() %>%
  layer_flatten(input_shape = c(120, num_features)) %>%
  layer_dense(units = 128, activation = "relu") %>%
  layer_dropout(rate = 0.30) %>%
  layer_dense(units = 64, activation = "relu") %>%
  layer_dropout(rate = 0.30) %>%
  layer_dense(units = 1, activation = "linear")

dnn_dropout %>% compile(
  optimizer = optimizer_adam(),
  loss = "mae",
  metrics = c("mae")
)

summary(dnn_dropout)

params_dropout <- count_trainable_params(dnn_dropout)
params_dropout

```

```{r}
history_dropout <- dnn_dropout %>% fit(
  train_gen,
  steps_per_epoch = steps_per_epoch,
  epochs = epochs,
  validation_data = val_gen,
  validation_steps = validation_steps,
  callbacks = list(early_stop)
)
```

# 3. DNN With L2 Weight Decay

```{r}
# =========================
# DNN MODEL 3: L2 REGULARIZATION
# =========================
dnn_l2 <- keras_model_sequential() %>%
  layer_flatten(input_shape = c(120, num_features)) %>%
  layer_dense(
    units = 128,
    activation = "relu",
    kernel_regularizer = regularizer_l2(0.001)
  ) %>%
  layer_dense(
    units = 64,
    activation = "relu",
    kernel_regularizer = regularizer_l2(0.001)
  ) %>%
  layer_dense(units = 1, activation = "linear")

dnn_l2 %>% compile(
  optimizer = optimizer_adam(),
  loss = "mae",
  metrics = c("mae")
)

summary(dnn_l2)

params_l2 <- count_trainable_params(dnn_l2)
params_l2
```

```{r}
history_l2 <- dnn_l2 %>% fit(
  train_gen,
  steps_per_epoch = steps_per_epoch,
  epochs = epochs,
  validation_data = val_gen,
  validation_steps = validation_steps,
  callbacks = list(early_stop)
)
```

## Parameter Count Table

```{r}
dnn_param_table <- data.frame(
  Model = c("DNN No Regularization", "DNN Dropout", "DNN L2 Regularization"),
  Trainable_Parameters = c(params_no_reg, params_dropout, params_l2)
)

dnn_param_table
```

# Learning Curves for the Three DNN Variants

```{r}
# =========================
# EXTRACT LEARNING CURVES
# =========================
extract_history <- function(history, model_name) {
  data.frame(
    Epoch = seq_along(history$metrics$mae),
    Training_MAE = history$metrics$mae,
    Validation_MAE = history$metrics$val_mae,
    Model = model_name
  )
}

hist_no_reg <- extract_history(history_no_reg, "No Regularization")
hist_dropout <- extract_history(history_dropout, "Dropout")
hist_l2 <- extract_history(history_l2, "L2 Regularization")

dnn_history_all <- bind_rows(hist_no_reg, hist_dropout, hist_l2)

dnn_history_long <- dnn_history_all %>%
  pivot_longer(
    cols = c(Training_MAE, Validation_MAE),
    names_to = "Curve",
    values_to = "MAE"
  )

ggplot(dnn_history_long, aes(x = Epoch, y = MAE, linetype = Curve)) +
  geom_line() +
  facet_wrap(~ Model, scales = "free_y") +
  labs(
    title = "Learning Curves for DNN Regression Variants",
    x = "Epoch",
    y = "Mean Absolute Error"
  ) +
  theme_minimal()
```

# **Training of DNN Regression Models**

The three DNN models employed for regression were trained using Adam as the optimiser and Mean Absolute Error (MAE) as the loss function. Mean absolute error was preferred because it gives the average difference between predicted and observed values, which can be measured in Celsius.

The early stopping technique was used, with patience of 5 epochs based on validation MAE. This implies that training will stop automatically if validation MAE is unable to become better for five consecutive epochs. The model weights will revert to their initial form before stopping to avoid keeping an overfitted model from later epochs.

There were three DNN models trained. The first one had no regularisation at all, and it was used as a baseline. Dropout was used in the second model by randomly deactivating some neurone connections. L2 weight decay was used in the third model by adding a penalty to the loss function to discourage excessive weights.

From the learning curves provided above, it is clear that the bias/variance nature of the different models can be analysed. If the first model with no regularisation is able to attain a small training MAE value but a much larger validation MAE, then it indicates that the model is overfitting and has a high variance. The other two models with regularisation are anticipated to perform well by minimising overfitting. However, too much regularisation causes high bias, resulting in more training and validation errors.

# SELECT BEST DNN VARIANT

```{r}
# =========================
# SELECT BEST DNN VARIANT
# =========================
best_val_mae <- c(
  No_Regularization = min(history_no_reg$metrics$val_mae),
  Dropout = min(history_dropout$metrics$val_mae),
  L2_Regularization = min(history_l2$metrics$val_mae)
)

best_val_mae

best_model_name <- names(which.min(best_val_mae))
best_model_name

best_dnn_model <- switch(
  best_model_name,
  "No_Regularization" = dnn_no_reg,
  "Dropout" = dnn_dropout,
  "L2_Regularization" = dnn_l2
)
```

# CREATE TEST ARRAYS FOR EVALUATION

```{r}
# =========================
# CREATE TEST ARRAYS FOR EVALUATION
# =========================
make_sequences <- function(data, lookback = 120, delay = 24) {
  
  data_values <- as.matrix(data %>% select(where(is.numeric)))
  max_index <- nrow(data_values) - delay
  
  rows <- (lookback + 1):max_index
  
  samples <- array(0, dim = c(length(rows), lookback, ncol(data_values)))
  targets <- numeric(length(rows))
  
  for (j in seq_along(rows)) {
    idx <- rows[j]
    indices <- (idx - lookback):(idx - 1)
    samples[j,,] <- data_values[indices,]
    targets[j] <- data$T..degC.[idx + delay]
  }
  
  list(samples = samples, targets = targets)
}

test_seq <- make_sequences(test_scaled, lookback = 120, delay = 24)

x_test_jena <- test_seq$samples
y_test_jena <- test_seq$targets

dim(x_test_jena)
length(y_test_jena)
```

# PREDICTIONS FROM BEST DNN MODEL

```{r}
# =========================
# PREDICTIONS FROM BEST DNN MODEL
# =========================
dnn_pred <- best_dnn_model %>%
  predict(x_test_jena)

dnn_pred <- as.numeric(dnn_pred)
```

# TEST METRICS

```{r}
# =========================
# TEST METRICS
# =========================
test_mae <- mean(abs(y_test_jena - dnn_pred))

test_rmse <- sqrt(mean((y_test_jena - dnn_pred)^2))

test_r2 <- 1 - sum((y_test_jena - dnn_pred)^2) /
  sum((y_test_jena - mean(y_test_jena))^2)

dnn_test_results <- data.frame(
  Best_Model = best_model_name,
  Test_MAE = round(test_mae, 4),
  Test_RMSE = round(test_rmse, 4),
  Test_R2 = round(test_r2, 4)
)

dnn_test_results
```

# Naive Baseline: Predict T+24 = T+0

```{r}
# =========================
# NAIVE BASELINE
# Predict future temperature using current temperature
# =========================
naive_pred <- numeric(length(y_test_jena))

for (j in seq_along(naive_pred)) {
  current_index <- 120 + j
  naive_pred[j] <- test_scaled$T..degC.[current_index]
}

naive_mae <- mean(abs(y_test_jena - naive_pred))

baseline_comparison <- data.frame(
  Model = c("Naive Baseline", paste("Best DNN:", best_model_name)),
  Test_MAE = c(round(naive_mae, 4), round(test_mae, 4))
)

baseline_comparison
```

# One-Week Predicted vs Actual Plot

```{r}
# =========================
# ONE WEEK TEST PREDICTIONS
# =========================
one_week_hours <- 24 * 7

plot_data_week <- data.frame(
  Time = test_data$Date.Hour[(120 + 24 + 1):(120 + 24 + one_week_hours)],
  Actual = y_test_jena[1:one_week_hours],
  Predicted = dnn_pred[1:one_week_hours]
)

plot_data_long <- plot_data_week %>%
  pivot_longer(cols = c(Actual, Predicted),
               names_to = "Series",
               values_to = "Temperature")

ggplot(plot_data_long, aes(x = Time, y = Temperature, linetype = Series)) +
  geom_line() +
  labs(
    title = "Predicted vs Actual Temperature for One Week of Test Data",
    x = "Date",
    y = "Temperature"
  ) +
  theme_minimal()
```

# **Test Evaluation of the Best DNN Variant**

The selected optimal DNN model will be based on the smallest MAE during validation. Such criteria ensure that the model chosen at the end of training relies only on unseen validation data and not training data alone.

On the test set, the selected optimal DNN model had an MAE of `r round(test_mae, 4)`, an RMSE of `r round(test_rmse, 4)` and an R squared of `r round(test_r2, 4)`. MAE represents the average difference between observed and predicted temperatures for one day ahead, while RMSE penalises differences with increasing weight for large differences. An R squared score of `round(test_r2, 4)` means that the model can explain 73% of the variability in temperatures 24 hours in advance.

The plot of weekly predictions for one week gives a graphical representation of how well the predictions follow the trends in actual temperature. Good predictions will capture the trend in rising and falling temperature, but short-term rapid changes in temperature can prove difficult.

The naive method used in this experiment predicts one-day ahead temperatures based on the current temperature level. Therefore, its MAE score of '0' indicates how much better the selected optimal DNN method can predict compared to a very simple baseline persistence method. If the best DNN model outperforms the baseline naive method, then the DNN model can make accurate predictions.

# **a. How a 2-D convolutional filter operates on a greyscale image versus how a 1-D filter operates on a multivariate time-series window**

A two-dimensional convolutional filter uses a greyscale image to work in two dimensions, namely height and width. The operation involves moving from one point to another in the image to apply the filter, where a 3×3 square is an example of a local neighbourhood of pixels whose pixel values are weighted together to create a feature map for detecting edges, curves, corners, and digit strokes.

A 1D filter does not operate like this; instead, it works in one dimension, which is time. It slides through the multivariate time series window to study consecutive local sections of time steps in the time series dataset. When the climate dataset of Jena is being analysed using hourly data, for example, a kernel size of 5 indicates five hours' worth of temporal patterns to identify short-term changes in temperature, humidity, and wind speeds.

# **b. Shared inductive bias and parameter efficiency**

The common inductive bias between two-dimensional and one-dimensional convolution is locality and parameter sharing. Locality implies that the network assumes proximity to be a stronger relationship than distance. For instance, adjacent pixels in an image create recognisable visual structures, while adjacent time points in a sequence tend to create recognisable temporal patterns.

On the other hand, parameter sharing implies that a single set of weights can be used multiple times on various sections of the data. Rather than having unique filters for each pixel or each data point, the CNN (Convolutional Neural Network) will have only one set of weights and apply it throughout the data. Consequently, the network requires fewer trainable parameters than a full-scale DNN.

# **c. Why local connectivity suits images and time series but is less natural for tabular data**

The idea of local connectivity works well for images, since neighbouring pixels generally represent one object or a similar pattern, such as strokes representing digits in handwriting. Thus, small filters can discover informative spatial features.

Time series benefit from the concept of local connectivity, as neighbouring elements often exhibit correlation. For instance, the temperature observed now is highly correlated with the temperature seen within the previous few hours. Thus, a one-dimensional filter can pick up short-term dependencies in data.

On the contrary, local connectivity seems inappropriate for working with tabular data, as tabular variables normally lack a natural order. For instance, pressure, humidity, wind speed, and temperature are likely to be related to each other, but the order they are positioned in a table does not necessarily mean they form a neighbourhood. Therefore, local filters cannot detect dependencies between these features when applied to regular tabular data.

# II.B(a) 2D-CNN MODEL FOR MNIST

```{r}
# =========================
# II.B(a) 2D-CNN MODEL FOR MNIST
# =========================

cnn_mnist <- keras_model_sequential() %>%
  layer_conv_2d(
    filters = 32,
    kernel_size = c(3, 3),
    activation = "relu",
    padding = "same",
    input_shape = c(28, 28, 1)
  ) %>%
  layer_conv_2d(
    filters = 64,
    kernel_size = c(3, 3),
    activation = "relu",
    padding = "same"
  ) %>%
  layer_max_pooling_2d(pool_size = c(2, 2)) %>%
  layer_flatten() %>%
  layer_dense(units = 128, activation = "relu") %>%
  layer_dense(units = 10, activation = "softmax")

summary(cnn_mnist)
```

# **B(a) 2D-CNN MODEL FOR MNIST**

The input of the 2D-CNN was set to `(28, 28, 1)`, denoting the input of 28×28 images that are greyscale MNIST with only one colour channel. The network consists of two Conv2D layers, followed by a max pooling layer and then another layer of convolution. All of these layers use ReLU activation and have the same padding. The convolutional layers are responsible for detecting local spatial features in the data, such as edges and digit strokes. The final output of the network is a softmax layer, which has 10 neurones.

# II.B(b) COMPILE CNN MODEL

```{r}
# =========================
# II.B(b) COMPILE CNN MODEL
# =========================

cnn_mnist %>% compile(
  optimizer = optimizer_adam(),
  loss = "categorical_crossentropy",
  metrics = c("accuracy")
)

early_stop_cnn <- callback_early_stopping(
  monitor = "val_loss",
  patience = 5,
  restore_best_weights = TRUE
)
```

# TRAINING TIME PER EPOCH

```{r}
# =========================
# TRAINING TIME PER EPOCH
# =========================

start_time <- Sys.time()

history_cnn_mnist <- cnn_mnist %>% fit(
  x_train_cnn, y_train_cat,
  epochs = 20,
  batch_size = 128,
  validation_data = list(x_val_cnn, y_val_cat),
  callbacks = list(early_stop_cnn)
)

end_time <- Sys.time()

cnn_total_time <- as.numeric(difftime(end_time, start_time, units = "secs"))
cnn_epochs_run <- length(history_cnn_mnist$metrics$loss)
cnn_time_per_epoch <- cnn_total_time / cnn_epochs_run

cnn_time_per_epoch  
```

# TEST ACCURACY

```{r}
# =========================
# TEST ACCURACY
# =========================

cnn_eval <- cnn_mnist %>% evaluate(x_test_cnn, y_test_cat)

cnn_test_loss <- cnn_eval["loss"]
cnn_test_accuracy <- cnn_eval["accuracy"]

cnn_test_accuracy
```

# PREDICTIONS AND CONFUSION MATRIX

```{r}
# =========================
# PREDICTIONS AND CONFUSION MATRIX
# =========================

cnn_prob <- cnn_mnist %>% predict(x_test_cnn)

cnn_pred_class <- apply(cnn_prob, 1, which.max) - 1

cnn_conf_matrix <- table(
  Actual = y_test_raw,
  Predicted = cnn_pred_class
)

cnn_conf_matrix
```

# MACRO-F1 FUNCTION

```{r}
# =========================
# MACRO-F1 FUNCTION
# =========================

macro_f1_score <- function(actual, predicted) {
  
  classes <- sort(unique(actual))
  
  f1_values <- sapply(classes, function(cls) {
    
    tp <- sum(actual == cls & predicted == cls)
    fp <- sum(actual != cls & predicted == cls)
    fn <- sum(actual == cls & predicted != cls)
    
    precision <- ifelse((tp + fp) == 0, 0, tp / (tp + fp))
    recall <- ifelse((tp + fn) == 0, 0, tp / (tp + fn))
    
    f1 <- ifelse((precision + recall) == 0, 0,
                 2 * precision * recall / (precision + recall))
    
    return(f1)
  })
  
  mean(f1_values)
}

cnn_macro_f1 <- macro_f1_score(y_test_raw, cnn_pred_class)

cnn_results <- data.frame(
  Model = "2D-CNN",
  Dataset = "MNIST",
  Test_Accuracy = round(as.numeric(cnn_test_accuracy), 4),
  Macro_F1 = round(cnn_macro_f1, 4),
  Train_Time_Per_Epoch_Seconds = round(cnn_time_per_epoch, 2)
)

cnn_results
```

# **B(b) COMPILE CNN MODEL**

The 2D CNN model was trained with the use of the Adam optimiser and categorical cross-entropy as the cost function, which is in line with the DNN classifier from Part I. Early stopping with patience 5 was employed as a training cessation mechanism once there were no further improvements in validation loss. The test accuracy and the macro F1 score for the 2D CNN model are '0.87' and '0.85', respectively.

# PARAMETER COUNT FUNCTION

```{r}
# =========================
# PARAMETER COUNT FUNCTION
# =========================

count_trainable_params <- function(model) {
  sum(sapply(model$trainable_weights, keras::k_count_params))
}

cnn_params <- count_trainable_params(cnn_mnist)

cnn_params
```

# II.C(a) 1D-CNN FOR JENA REGRESSION

```{r}
# =========================
# II.C(a) 1D-CNN FOR JENA REGRESSION
# =========================

cnn1d_jena <- keras_model_sequential() %>%
  layer_conv_1d(
    filters = 32,
    kernel_size = 5,
    activation = "relu",
    padding = "same",
    input_shape = c(120, num_features)
  ) %>%
  layer_conv_1d(
    filters = 64,
    kernel_size = 5,
    activation = "relu",
    padding = "same"
  ) %>%
  layer_max_pooling_1d(pool_size = 2) %>%
  layer_global_average_pooling_1d() %>%
  layer_dense(units = 64, activation = "relu") %>%
  layer_dense(units = 1, activation = "linear")

summary(cnn1d_jena)
```

# **C(a) 1D-CNN FOR JENA REGRESSION**

The architecture for the 1D-CNN model consists of Conv1D layers having ReLU activation with an input shape of `(120, no of predictors)` and one MaxPooling1D, GlobalAveragePooling1D, and Linear layers. Input shape (120, number of predictors) implies that there are 120 previous data points corresponding to the last 120 hours, whereas the other dimension corresponds to the predictors of the meteorological data. It is a regression problem, and thus, the output should be linear.

# **b.Justify kernel size and temporal receptive field**

Conv1D layers have a kernel size of 5. Given that the Jena dataset has been resampled from one hour to another, a kernel size of 5 refers to a local temporal window of five hours in which each convolutional filter operates. Such convolution will enable the neural network to learn short-range features, such as temperature rising gradually, periods of cooling, humidity, and pressure changes.

Using two Conv1D layers means that the model has an enhanced temporal receptive field. In other words, while the first layer learns local features of the five-hour window, the second layer learns features formed by combining those local five-hour windows together. Consequently, even though each kernel is operating for five hours, the two-layer stack enables learning more complex temporal features within a larger span of the 120-hour window.

A kernel size of 5 is sufficient for the purpose of capturing short-range weather variation without over-smoothing the input sequence of values.

# II.C(c) COMPILE 1D-CNN

```{r}
# =========================
# II.C(c) COMPILE 1D-CNN
# =========================

cnn1d_jena %>% compile(
  optimizer = optimizer_adam(),
  loss = "mae",
  metrics = c("mae")
)

early_stop_cnn1d <- callback_early_stopping(
  monitor = "val_mae",
  patience = 5,
  restore_best_weights = TRUE
)
```

# TRAIN 1D-CNN

```{r}
# =========================
# TRAIN 1D-CNN
# =========================

start_time <- Sys.time()

history_cnn1d_jena <- cnn1d_jena %>% fit(
  train_gen,
  steps_per_epoch = steps_per_epoch,
  epochs = epochs,
  validation_data = val_gen,
  validation_steps = validation_steps,
  callbacks = list(early_stop_cnn1d)
)

end_time <- Sys.time()

cnn1d_total_time <- as.numeric(difftime(end_time, start_time, units = "secs"))
cnn1d_epochs_run <- length(history_cnn1d_jena$metrics$loss)
cnn1d_time_per_epoch <- cnn1d_total_time / cnn1d_epochs_run
```

# TEST EVALUATION

```{r}
# =========================
# TEST EVALUATION
# =========================

cnn1d_pred <- cnn1d_jena %>% predict(x_test_jena)
cnn1d_pred <- as.numeric(cnn1d_pred)

cnn1d_mae <- mean(abs(y_test_jena - cnn1d_pred))

cnn1d_rmse <- sqrt(mean((y_test_jena - cnn1d_pred)^2))

cnn1d_r2 <- 1 - sum((y_test_jena - cnn1d_pred)^2) /
  sum((y_test_jena - mean(y_test_jena))^2)

cnn1d_params <- count_trainable_params(cnn1d_jena)

cnn1d_results <- data.frame(
  Model = "1D-CNN",
  Test_MAE = round(cnn1d_mae, 4),
  Test_RMSE = round(cnn1d_rmse, 4),
  Test_R2 = round(cnn1d_r2, 4),
  Parameters = cnn1d_params,
  Train_Time_Per_Epoch_Seconds = round(cnn1d_time_per_epoch, 2)
)

cnn1d_results
```

# COMPARE 1D-CNN WITH DNN BASELINE

```{r}
# =========================
# COMPARE 1D-CNN WITH DNN BASELINE
# =========================

jena_dnn_cnn_comparison <- data.frame(
  Model = c("DNN Baseline", "1D-CNN"),
  Test_MAE = c(round(test_mae, 4), round(cnn1d_mae, 4)),
  Test_RMSE = c(round(test_rmse, 4), round(cnn1d_rmse, 4)),
  Test_R2 = c(round(test_r2, 4), round(cnn1d_r2, 4)),
  Parameters = c(
    ifelse(exists("params_no_reg"), params_no_reg, NA),
    cnn1d_params
  ),
  Train_Time_Per_Epoch_Seconds = c(
    ifelse(exists("dnn_reg_time_per_epoch"), round(dnn_reg_time_per_epoch, 2), NA),
    round(cnn1d_time_per_epoch, 2)
  )
)

jena_dnn_cnn_comparison
```

# **COMPARE 1D-CNN WITH DNN BASELINE**

The Adam optimiser and MAE loss were utilised for training the 1D-CNN, identical to the DNN regression baseline approach. On the Jena dataset, the MAE was `r round(cnn1d_mae, 4)`, the RMSE was `r round(cnn1d_rmse, 4)`, and the R² was `r round(cnn1d_r2, 4)`.

When compared to the DNN regression baseline, the 1D-CNN maintains the temporal dynamics within the input sequence of 120 hours. The DNN model converts the time-series inputs into a flattened vector and considers all lagged observations as regular features. Conversely, the 1D-CNN model applies filters on neighbouring hour observations. Thus, the 1D-CNN model should be superior in identifying local temporal relations. A lower MAE and RMSE or higher R² for the 1D-CNN over the DNN baseline implies the advantage of exploiting temporal locality for climate forecasting.

# III.A Theoretical Motivation

# **a. Why a vanilla SimpleRNN struggles with sequences of 120+ steps**

It becomes difficult to pass data over a long period using a vanilla simple RNN since the data passes through the same recurrent operation over and over again. The issue is that while using backpropagation through time, there is an accumulation of repeated multiplication between the gradients and the recurrent matrices. When these multiplications are less than one, the resulting gradients become negligible over several time periods.

# **b. How LSTM gates carry information over long horizons**

The Long Short-Term Memory architecture solves the problem of gradient disappearance through the addition of the cell state and three basic gates, namely, the forget gate, the input gate, and the output gate.

The forget gate determines what part of the information from the previous cell state should be stored and what should be forgotten. The input gate determines what new data should be included in the cell state. Finally, the output gate regulates how much data from the cell state should be sent to the hidden state for predictions.

The cell state can be viewed as a path that enables information transmission across multiple time points with less frequent nonlinear transformations. As a result, the LSTM network finds it much simpler to retain relevant data for an extended period of time. It is especially helpful in the case of the Jena climate data set since the temperature after 24 hours depends on both current and previous temperatures during 120 hours of data look-back.

# **c. Why treating MNIST rows as time steps is valid but suboptimal**

The treatment of the MNIST images as a sequence is appropriate because each image could be described as a 28-row sequence with 28 pixel values per row. With this representation, the LSTM network would sequentially process the image, row by row, in 28 time steps with 28 features per step.

Nevertheless, such a usage of LSTM is inappropriate, as the task of classifying the handwritten digits based on the spatial image is more appropriate for CNN than LSTM networks. This limitation is due to the spatial structure of the pixel arrangement, which includes 2D strokes, curves, and edges. The CNN, being capable of learning the filter movements over the height and width of the image, is able to extract such spatial features.

Therefore, using an LSTM for solving MNIST classification demonstrates the significance of inductive bias. Although there is sufficient capacity to perform the digit recognition task, the sequential nature of the LSTM biases will not be as beneficial as the spatial locality and parameter sharing of CNN networks.

# III.B(a) LSTM MODEL FOR JENA REGRESSION

```{r}
# =========================
# III.B(a) LSTM MODEL FOR JENA REGRESSION
# =========================

lstm_jena <- keras_model_sequential() %>%
  
  # First LSTM layer (returns sequences)
  layer_lstm(
    units = 64,
    return_sequences = TRUE,
    input_shape = c(120, num_features),
    recurrent_dropout = 0.2
  ) %>%
  
  # Second LSTM layer
  layer_lstm(
    units = 32,
    return_sequences = FALSE,
    recurrent_dropout = 0.2
  ) %>%
  
  # Dense output
  layer_dense(units = 1, activation = "linear")

summary(lstm_jena)
```

# **The LSTM model was designed**

The LSTM was constructed to receive inputs of size (120, number of predictors) 120, where 120 hours represented the lookback period. Two LSTM layers were stacked, in which the first one returned sequences such that the second LSTM could fully analyse the temporal data. Dropout was applied during training on the recurrent units to avoid overfitting of the data. Finally, a linear activation function was applied in the output layer due to the fact that this was a regression problem.

# b. Train and Evaluate LSTM(COMPILE LSTM MODEL)

```{r}
# =========================
# COMPILE LSTM MODEL
# =========================
lstm_jena %>% compile(
  optimizer = optimizer_adam(),
  loss = "mae",
  metrics = c("mae")
)

early_stop_lstm <- callback_early_stopping(
  monitor = "val_mae",
  patience = 5,
  restore_best_weights = TRUE
)
```

# TRAIN LSTM

```{r}
# =========================
# TRAIN LSTM
# =========================

start_time <- Sys.time()

history_lstm_jena <- lstm_jena %>% fit(
  train_gen,
  steps_per_epoch = steps_per_epoch,
  epochs = epochs,
  validation_data = val_gen,
  validation_steps = validation_steps,
  callbacks = list(early_stop_lstm)
)

end_time <- Sys.time()

lstm_total_time <- as.numeric(difftime(end_time, start_time, units = "secs"))
lstm_epochs_run <- length(history_lstm_jena$metrics$loss)
lstm_time_per_epoch <- lstm_total_time / lstm_epochs_run
```

# TEST EVALUATION

```{r}
# =========================
# TEST EVALUATION
# =========================

lstm_pred <- lstm_jena %>% predict(x_test_jena)
lstm_pred <- as.numeric(lstm_pred)

lstm_mae <- mean(abs(y_test_jena - lstm_pred))

lstm_rmse <- sqrt(mean((y_test_jena - lstm_pred)^2))

lstm_r2 <- 1 - sum((y_test_jena - lstm_pred)^2) /
  sum((y_test_jena - mean(y_test_jena))^2)

lstm_params <- count_trainable_params(lstm_jena)

lstm_results <- data.frame(
  Model = "LSTM",
  Test_MAE = round(lstm_mae, 4),
  Test_RMSE = round(lstm_rmse, 4),
  Test_R2 = round(lstm_r2, 4),
  Parameters = lstm_params,
  Train_Time_Per_Epoch_Seconds = round(lstm_time_per_epoch, 2)
)

lstm_results
```

# c. Ablation: SimpleRNN vs LSTM

```{r}
# =========================
# SIMPLE RNN MODEL (ABLATION)
# =========================

rnn_jena <- keras_model_sequential() %>%
  layer_simple_rnn(
    units = 64,
    return_sequences = TRUE,
    input_shape = c(120, num_features)
  ) %>%
  layer_simple_rnn(units = 32) %>%
  layer_dense(units = 1, activation = "linear")

rnn_jena %>% compile(
  optimizer = optimizer_adam(),
  loss = "mae",
  metrics = c("mae")
)
```

# TRAIN SIMPLE RNN

```{r}
# TRAIN SIMPLE RNN

history_rnn_jena <- rnn_jena %>% fit(
  train_gen,
  steps_per_epoch = steps_per_epoch,
  epochs = epochs,
  validation_data = val_gen,
  validation_steps = validation_steps,
  callbacks = list(early_stop_lstm)
)
```

# TEST PERFORMANCE OF SIMPLE RNN

```{r}
# TEST PERFORMANCE OF SIMPLE RNN

rnn_pred <- rnn_jena %>% predict(x_test_jena)
rnn_pred <- as.numeric(rnn_pred)

rnn_mae <- mean(abs(y_test_jena - rnn_pred))

rnn_mae
```

# **LSTM vs SimpleRNN for Jena Regression**

The LSTM network was also trained on the same dataset with the same loss function as well as stopping criterion as used in the DNN model. During testing, the LSTM model obtained an MAE score of `r round(lstm_mae, 4)`, an RMSE of `r round(lstm_rmse, 4)` and an R² of `r round(lstm_r2, 4)`.

In order to analyse the significance of long-term memory, an ablation study was conducted, where a SimpleRNN with the same architecture was trained. The MAE score obtained by the SimpleRNN on the test set was '0.001'.

This difference in accuracy between the LSTM and the SimpleRNN can be attributed to the vanishing gradient problem faced by the latter. While the SimpleRNN model struggles to learn long-range correlations in the 120-hour input sequence, due to the vanishing gradients problem, the LSTM network uses gating mechanisms and its cell state to remember information. This allows the LSTM model to forecast more accurately than the SimpleRNN.

# IV a. Create classification generator and LSTM model

```{r}
# =========================
# PREPARE JENA CLASSIFICATION DATA
# =========================

jena_class_data <- jena_hourly %>%
  arrange(Date.Hour) %>%
  mutate(
    temp_24h_ahead = lead(T..degC., 24),
    temp_class = case_when(
      temp_24h_ahead < 0 ~ "Cold",
      temp_24h_ahead >= 0 & temp_24h_ahead < 10 ~ "Cool",
      temp_24h_ahead >= 10 & temp_24h_ahead < 20 ~ "Mild",
      temp_24h_ahead >= 20 ~ "Warm"
    )
  ) %>%
  filter(!is.na(temp_24h_ahead), !is.na(temp_class))

train_class <- jena_class_data %>%
  filter(Date.Hour >= ymd_hms("2009-01-01 00:00:00"),
         Date.Hour <  ymd_hms("2015-01-01 00:00:00"))

val_class <- jena_class_data %>%
  filter(Date.Hour >= ymd_hms("2015-01-01 00:00:00"),
         Date.Hour <  ymd_hms("2016-01-01 00:00:00"))

test_class <- jena_class_data %>%
  filter(Date.Hour >= ymd_hms("2016-01-01 00:00:00"),
         Date.Hour <  ymd_hms("2017-01-01 00:00:00"))
```

# CLASSIFICATION DATA GENERATOR

```{r}
# =========================
# CLASSIFICATION DATA GENERATOR
# =========================

jena_class_generator <- function(data,
                                 lookback = 120,
                                 delay = 24,
                                 step = 1,
                                 batch_size = 128,
                                 shuffle = FALSE) {
  
  class_levels <- c("Cold", "Cool", "Mild", "Warm")
  
  predictor_cols <- names(data %>%
                            select(where(is.numeric))) 
  
  predictor_cols <- setdiff(predictor_cols, c("temp_24h_ahead"))
  
  data_values <- as.matrix(data[, predictor_cols])
  
  max_index <- nrow(data_values) - delay - 1
  
  function() {
    
    if (shuffle) {
      rows <- sample((lookback + 1):max_index, batch_size)
    } else {
      rows <- seq(lookback + 1, min(lookback + batch_size, max_index))
    }
    
    samples <- array(0, dim = c(length(rows), lookback, ncol(data_values)))
    targets <- matrix(0, nrow = length(rows), ncol = 4)
    
    for (j in seq_along(rows)) {
      idx <- rows[j]
      indices <- seq(idx - lookback, idx - 1, by = step)
      
      samples[j,,] <- data_values[indices,]
      
      class_id <- match(data$temp_class[idx + delay], class_levels)
      targets[j, class_id] <- 1
    }
    
    list(samples, targets)
  }
}
```

# CREATE CLASSIFICATION GENERATORS

```{r}
# =========================
# CREATE CLASSIFICATION GENERATORS
# =========================

train_class_gen <- jena_class_generator(
  train_class,
  batch_size = 128,
  shuffle = TRUE
)

val_class_gen <- jena_class_generator(
  val_class,
  batch_size = 128,
  shuffle = FALSE
)

test_class_gen <- jena_class_generator(
  test_class,
  batch_size = 128,
  shuffle = FALSE
)

# Verify shape
class_batch <- train_class_gen()

dim(class_batch[[1]])
dim(class_batch[[2]])
```

# LSTM MODEL FOR JENA CLASSIFICATION

```{r}
# =========================
# LSTM MODEL FOR JENA CLASSIFICATION
# =========================

num_features_class <- dim(class_batch[[1]])[3]

lstm_jena_class <- keras_model_sequential() %>%
  layer_lstm(
    units = 64,
    return_sequences = TRUE,
    input_shape = c(120, num_features_class),
    recurrent_dropout = 0.2
  ) %>%
  layer_lstm(
    units = 32,
    recurrent_dropout = 0.2
  ) %>%
  layer_dense(units = 4, activation = "softmax")

summary(lstm_jena_class)
```

# **Jena classification task**

In order to solve the Jena classification problem, a 4-neurone output layer was employed as part of the softmax activation instead of using the regression output layer. These 4 output neurones represent each class of temperature, namely cold, cool, mild, and warm. A softmax activation function works well in this case since the resulting class predictions sum up to one. The loss function was adjusted from mean absolute error to categorical cross-entropy.

# b. Train and evaluate: accuracy, macro-F1, confusion matrix

```{r}
# =========================
# COMPILE LSTM CLASSIFICATION MODEL
# =========================

lstm_jena_class %>% compile(
  optimizer = optimizer_adam(),
  loss = "categorical_crossentropy",
  metrics = c("accuracy")
)

early_stop_lstm_class <- callback_early_stopping(
  monitor = "val_loss",
  patience = 5,
  restore_best_weights = TRUE
)
```

# TRAIN LSTM CLASSIFICATION MODEL

```{r}
# =========================
# TRAIN LSTM CLASSIFICATION MODEL
# =========================

history_lstm_jena_class <- lstm_jena_class %>% fit(
  train_class_gen,
  steps_per_epoch = steps_per_epoch,
  epochs = epochs,
  validation_data = val_class_gen,
  validation_steps = validation_steps,
  callbacks = list(early_stop_lstm_class)
)
```

# CREATE TEST CLASSIFICATION ARRAYS

```{r}
# =========================
# CREATE TEST CLASSIFICATION ARRAYS
# =========================

make_class_sequences <- function(data, lookback = 120, delay = 24) {
  
  class_levels <- c("Cold", "Cool", "Mild", "Warm")
  
  predictor_cols <- names(data %>% select(where(is.numeric)))
  predictor_cols <- setdiff(predictor_cols, c("temp_24h_ahead"))
  
  data_values <- as.matrix(data[, predictor_cols])
  max_index <- nrow(data_values) - delay
  
  rows <- (lookback + 1):max_index
  
  samples <- array(0, dim = c(length(rows), lookback, ncol(data_values)))
  targets <- matrix(0, nrow = length(rows), ncol = 4)
  
  for (j in seq_along(rows)) {
    idx <- rows[j]
    samples[j,,] <- data_values[(idx - lookback):(idx - 1),]
    
    class_id <- match(data$temp_class[idx + delay], class_levels)
    targets[j, class_id] <- 1
  }
  
  list(samples = samples, targets = targets)
}

test_class_seq <- make_class_sequences(test_class)

x_test_jena_class <- test_class_seq$samples
y_test_jena_class <- test_class_seq$targets
```

# TEST PREDICTION

```{r}
# =========================
# TEST PREDICTION
# =========================

lstm_class_prob <- lstm_jena_class %>% predict(x_test_jena_class)

lstm_class_pred <- apply(lstm_class_prob, 1, which.max)

lstm_class_actual <- apply(y_test_jena_class, 1, which.max)

class_levels <- c("Cold", "Cool", "Mild", "Warm")

lstm_class_pred_label <- factor(class_levels[lstm_class_pred], levels = class_levels)
lstm_class_actual_label <- factor(class_levels[lstm_class_actual], levels = class_levels)
```

# ACCURACY, MACRO-F1, CONFUSION MATRIX

```{r}
# =========================
# ACCURACY, MACRO-F1, CONFUSION MATRIX
# =========================

lstm_class_accuracy <- mean(lstm_class_pred_label == lstm_class_actual_label)

macro_f1_score <- function(actual, predicted) {
  
  classes <- levels(actual)
  
  f1_values <- sapply(classes, function(cls) {
    
    tp <- sum(actual == cls & predicted == cls)
    fp <- sum(actual != cls & predicted == cls)
    fn <- sum(actual == cls & predicted != cls)
    
    precision <- ifelse((tp + fp) == 0, 0, tp / (tp + fp))
    recall <- ifelse((tp + fn) == 0, 0, tp / (tp + fn))
    
    ifelse((precision + recall) == 0, 0,
           2 * precision * recall / (precision + recall))
  })
  
  mean(f1_values)
}

lstm_class_macro_f1 <- macro_f1_score(
  lstm_class_actual_label,
  lstm_class_pred_label
)

lstm_class_conf_matrix <- table(
  Actual = lstm_class_actual_label,
  Predicted = lstm_class_pred_label
)

lstm_class_results <- data.frame(
  Model = "LSTM",
  Dataset = "Jena 4-Class",
  Test_Accuracy = round(lstm_class_accuracy, 4),
  Macro_F1 = round(lstm_class_macro_f1, 4)
)

lstm_class_results
lstm_class_conf_matrix
```

# **The LSTM classification model**

The accuracy attained by the LSTM classifier on testing is `r round(lstm_class_accuracy, 4)` while the macro F1-score is `r round(lstm_class_macro_f1, 4)` . The confusion matrix gives the distribution of classification results for the four temperature classes. Misclassification should happen between adjacent temperature classes, e.g., cool and mild or mild and warm, since their temperature ranges overlap.

# **c. Discussion: Regression or classification advantage?**

As compared to the derived classification problem, the LSTM network will be more advantageous for the Jena regression prediction. The reason is that the LSTM network can take advantage of the continuous temporal dependencies, day-night cycles, and even multiple-day weather patterns in the training data due to its ability to predict the 24-hour-ahead temperature value exactly.

In the case of classification, the continuous value of the target is discretised into four categories. In doing so, part of the information is lost since the temperatures that are close to each other but on the border of two adjacent classes are put into two different categories. For instance, 9.9°C corresponds to the category 'Cool', while 10.0°C belongs to the category 'Mild'.

In theory, the LSTM networks are designed to model sequential dependence thanks to their gating mechanism and hence have an advantage in tasks requiring long-term temporal information extraction. Nevertheless, upon discretisation, some part of the sequential information might get lost, and thus, the LSTM network would be advantageous only for regression, limiting its effectiveness in classification tasks where temporal patterns are crucial.

# III.D LSTM for MNIST Sequential Image

# a. Reshape MNIST images to (28, 28)

```{r}
# =========================
# III.D(a) RESHAPE MNIST FOR LSTM
# =========================

# Each image is treated as 28 time steps,
# with 28 pixel values as features at each time step.

x_train_lstm_mnist <- array_reshape(x_train_final, c(nrow(x_train_final), 28, 28))
x_val_lstm_mnist   <- array_reshape(x_val, c(nrow(x_val), 28, 28))
x_test_lstm_mnist  <- array_reshape(x_test, c(nrow(x_test), 28, 28))

# Verify dimensions
dim(x_train_lstm_mnist)
dim(x_val_lstm_mnist)
dim(x_test_lstm_mnist)
```

# **Reshape MNIST images to (28, 28)**

MNIST images were reshaped such that they have 28 time steps, whereby each of the 28 time steps represents a row within the image that has 28 pixel features. This means that each image has an input shape of `(28, 28)`.

# b. Design and train LSTM model

```{r}
# =========================
# III.D(b) LSTM MODEL FOR MNIST
# =========================

lstm_mnist <- keras_model_sequential() %>%
  layer_lstm(
    units = 64,
    input_shape = c(28, 28),
    recurrent_dropout = 0.2
  ) %>%
  layer_dense(units = 128, activation = "relu") %>%
  layer_dense(units = 10, activation = "softmax")

summary(lstm_mnist)
```

# COMPILE MNIST LSTM

```{r}
# =========================
# COMPILE MNIST LSTM
# =========================

lstm_mnist %>% compile(
  optimizer = optimizer_adam(),
  loss = "categorical_crossentropy",
  metrics = c("accuracy")
)

early_stop_lstm_mnist <- callback_early_stopping(
  monitor = "val_loss",
  patience = 5,
  restore_best_weights = TRUE
)
```

# TRAIN MNIST LSTM

```{r}
# =========================
# TRAIN MNIST LSTM
# =========================

start_time <- Sys.time()

history_lstm_mnist <- lstm_mnist %>% fit(
  x_train_lstm_mnist,
  y_train_cat,
  epochs = 20,
  batch_size = 128,
  validation_data = list(x_val_lstm_mnist, y_val_cat),
  callbacks = list(early_stop_lstm_mnist)
)

end_time <- Sys.time()

lstm_mnist_total_time <- as.numeric(difftime(end_time, start_time, units = "secs"))
lstm_mnist_epochs_run <- length(history_lstm_mnist$metrics$loss)
lstm_mnist_time_per_epoch <- lstm_mnist_total_time / lstm_mnist_epochs_run
```

# c. Test accuracy and comparison with CNN

```{r}
# =========================
# TEST EVALUATION
# =========================

lstm_mnist_eval <- lstm_mnist %>% evaluate(
  x_test_lstm_mnist,
  y_test_cat
)

lstm_mnist_test_accuracy <- lstm_mnist_eval["accuracy"]

lstm_mnist_test_accuracy
```

# PARAMETER COUNT

```{r}
# =========================
# PARAMETER COUNT
# =========================

lstm_mnist_params <- count_trainable_params(lstm_mnist)

lstm_mnist_params
```

# COMPARE LSTM WITH CNN

```{r}
# =========================
# COMPARE LSTM WITH CNN
# =========================

mnist_lstm_cnn_comparison <- data.frame(
  Model = c("2D-CNN", "LSTM Sequential Image"),
  Test_Accuracy = c(
    round(as.numeric(cnn_test_accuracy), 4),
    round(as.numeric(lstm_mnist_test_accuracy), 4)
  ),
  Parameters = c(
    cnn_params,
    lstm_mnist_params
  ),
  Train_Time_Per_Epoch_Seconds = c(
    round(cnn_time_per_epoch, 2),
    round(lstm_mnist_time_per_epoch, 2)
  )
)

mnist_lstm_cnn_comparison
```

# **The MNIST LSTM model**

The MNIST LSTM model was able to get an accuracy of `0.98` on the test set. For the MNIST LSTM model, each image was considered as a sequence of 28 rows, where each row has 28 pixel features.

Since the MNIST handwritten digit classification task involves pattern recognition based on spatial attributes of the images, we expect that the LSTM will be inferior to the 2D-CNN since it treats each input sample as sequences of data points rather than a multidimensional tensor. In particular, the 2D-CNN architecture allows extracting information about local spatial features such as edges and curves because the filters scan the image in two dimensions.

It can be concluded that the accuracy gap between the CNN and the LSTM is due to different types of inductive biases. While the LSTM might have sufficient capacity to classify the digits, the inductive bias used for learning tasks from sequential data is suboptimal in the case of image processing.

# **Part IV.1 Design Brief — Custom Hybrid Architecture**

The chosen custom hybrid architecture is the CNN-LSTM model for solving the temperature regression issue in Jena. The custom hybrid architecture includes a 1D convolutional neural network along with the LSTM model. The reasons motivating the choice of this architecture lie in the weaknesses of the respective standard models.

Specifically, the DNN model uses a fully connected structure which considers all observations in the input dataset as features for regression in the 120-hour period. In other words, such architecture breaks the order of climate observations, which makes it unable to identify local time-dependent relationships. The 1D-CNN (one-dimensional convolutional neural network) solves this issue by detecting short-term time-dependent patterns such as variations of temperature, humidity, and air pressure during one hour. At the same time, this architecture fails to identify long-term dependencies in climate data. The ability to do so belongs to the LSTM networks. On the other hand, they may become computationally expensive and be less efficient in extracting local features due to their nature, which can hinder their performance in real-time applications where quick processing is essential.

The hybrid CNN-LSTM architecture is designed to overcome these issues. First, Conv1D layers enable the extraction of local temporal features out of the multivariate climate sequences. Then, the features obtained are passed to the LSTM layer for building the dependencies between the elements in this sequence. As a result, the custom hybrid architecture incorporates two inductive biases: the extraction of local features through convolution and longer sequential relationships identification through LSTM.

The chosen problem domain of weather forecasting is a good match for the suggested hybrid architecture because weather data include different patterns of temporal behaviour, both short-term ones taking place over a few hours and longer daily patterns covering a 120-hour period. Since the target variable of regression is the future 24-hour temperature value, a model that considers both local changes in the climate data and its long-term dependencies would work properly.

# **IV.1 Design Brief (Using Suggested Hybrid Pattern: CNN–LSTM)**

A hybrid architecture used in this work is CNN-LSTM, which consists of one-dimensional convolutional layers together with LSTM layers to solve Jena's temperature regression problem. The reason for choosing such architecture lies in the shortcomings of pure models like DNN, CNN, or LSTM.

First of all, the DNN-based model disregards the temporal nature of the data, as it uses the flattened representation of the input data. The second model, 1D-CNN, performs better compared to DNN due to its ability to use filters for capturing the local temporal dependencies in the data. However, it cannot model dependencies between observations located far from each other in the sequence of the 120-hour lookback window. Contrarily, LSTM is able to learn long-range dependencies due to its gating function; however, it might be less efficient at extracting short-range local patterns than convolutions, which can lead to challenges in accurately modelling data that requires both short and long-range context.

The combination of CNN and LSTM resolves both shortcomings of models mentioned above, as the CNN-based feature extraction layer extracts local temporal features such as local weather changes (hourly, daily, or week-based). Thereafter, the sequence of features is passed as input data to an LSTM-based layer, which learns long-term dependencies and temporal relationships. Consequently, such a network learns two inductive biases: local features using CNN and long-range memory using LSTM.

The Jena climate dataset is well-suited for testing of the described approach since the weather is characterised by periodicity in the form of weather changes. Moreover, weather changes can also occur locally within a period. Hence, since we aim to make a forecast in the next 24 hours based on 120-hour history, our architecture is suitable for such a task.

# IV.2 CNN-LSTM HYBRID MODEL USING FUNCTIONAL API

```{r}
# =========================
# IV.2 CNN-LSTM HYBRID MODEL USING FUNCTIONAL API
# =========================

hybrid_input <- layer_input(shape = c(120, num_features), name = "Input_120_hours")

hybrid_output <- hybrid_input %>%
  layer_conv_1d(
    filters = 32,
    kernel_size = 5,
    activation = "relu",
    padding = "same",
    name = "Conv1D_Local_Features_1"
  ) %>%
  layer_conv_1d(
    filters = 64,
    kernel_size = 5,
    activation = "relu",
    padding = "same",
    name = "Conv1D_Local_Features_2"
  ) %>%
  layer_max_pooling_1d(
    pool_size = 2,
    name = "MaxPooling_Temporal_Reduction"
  ) %>%
  layer_lstm(
    units = 64,
    recurrent_dropout = 0.2,
    name = "LSTM_Long_Range_Dependencies"
  ) %>%
  layer_dense(
    units = 64,
    activation = "relu",
    name = "Dense_Feature_Integration"
  ) %>%
  layer_dense(
    units = 1,
    activation = "linear",
    name = "Temperature_Output"
  )

hybrid_cnn_lstm <- keras_model(
  inputs = hybrid_input,
  outputs = hybrid_output,
  name = "CNN_LSTM_Hybrid_Jena_Regression"
)

summary(hybrid_cnn_lstm)
```

# COMPILE HYBRID MODEL

```{r}
# =========================
# COMPILE HYBRID MODEL
# =========================

hybrid_cnn_lstm %>% compile(
  optimizer = optimizer_adam(),
  loss = "mae",
  metrics = c("mae")
)

early_stop_hybrid <- callback_early_stopping(
  monitor = "val_mae",
  patience = 5,
  restore_best_weights = TRUE
)
```

# TRAIN HYBRID MODEL

```{r}
# =========================
# TRAIN HYBRID MODEL
# =========================

start_time <- Sys.time()

history_hybrid <- hybrid_cnn_lstm %>% fit(
  train_gen,
  steps_per_epoch = steps_per_epoch,
  epochs = epochs,
  validation_data = val_gen,
  validation_steps = validation_steps,
  callbacks = list(early_stop_hybrid)
)

end_time <- Sys.time()

hybrid_total_time <- as.numeric(difftime(end_time, start_time, units = "secs"))
hybrid_epochs_run <- length(history_hybrid$metrics$loss)
hybrid_time_per_epoch <- hybrid_total_time / hybrid_epochs_run
```

# **Architecture Diagram (CNN-LSTM Hybrid Architecture Diagram)**

Input Layer\
`(120, number of predictors)`\
⬇\
Conv1D Layer 1\
`32 filters, kernel size = 5, ReLU, same padding`\
⬇\
Conv1D Layer 2\
`64 filters, kernel size = 5, ReLU, same padding`\
⬇\
MaxPooling1D\
`pool size = 2`\
⬇\
LSTM Layer\
`64 units, recurrent dropout = 0.20`\
⬇\
Dense Layer\
`64 neurons, ReLU`\
⬇\
Output Layer\
`1 neuron, linear activation`

# **Implementation of the CNN-LSTM Hybrid Architecture**

The CNN-LSTM hybrid model was built through the functional API provided by Keras. It fulfils the requirement of designing a hybrid model architecture which goes beyond the traditional sequential model architecture and is created explicitly using `layer_input`and layer connections.

The model receives as input a tensor with dimensions `(120, number of predictors)`, corresponding to the previous 120 hours of Jena climate variables time series. The first two Conv1D layers extract local features in the temporal domain of the input sequence. The selected kernel size is 5 because the filters will learn features corresponding to short five-hour weather patterns. After applying the Conv1D layers, the temporal dimension is downsampled, and only relevant local information remains through the MaxPooling1D layer.

The generated sequence of features is fed into an LSTM layer with 64 units, which can learn temporal dependencies after extracting the short-term patterns. Finally, the sequence of features is aggregated by a dense layer and outputted by a single linear neurone which performs the prediction of the 24-hour-ahead temperature.

The model was trained through the same chronologically defined data splits and early stopping criteria applied to the other Jena models using the MAE loss function.

# IV.3 HYPERPARAMETER: NUMBER OF FILTERS

```{r}
# =========================
# IV.3 HYPERPARAMETER: NUMBER OF FILTERS
# =========================

filter_values <- c(16, 32, 64)

val_mae_results <- c()

for (f in filter_values) {
  
  model <- keras_model(
    inputs = hybrid_input,
    outputs =
      hybrid_input %>%
      layer_conv_1d(filters = f, kernel_size = 5, activation = "relu", padding = "same") %>%
      layer_conv_1d(filters = f * 2, kernel_size = 5, activation = "relu", padding = "same") %>%
      layer_max_pooling_1d(pool_size = 2) %>%
      layer_lstm(units = 64, recurrent_dropout = 0.2) %>%
      layer_dense(units = 64, activation = "relu") %>%
      layer_dense(units = 1, activation = "linear")
  )
  
  model %>% compile(
    optimizer = optimizer_adam(),
    loss = "mae",
    metrics = c("mae")
  )
  
  history <- model %>% fit(
    train_gen,
    steps_per_epoch = steps_per_epoch,
    epochs = epochs,
    validation_data = val_gen,
    validation_steps = validation_steps,
    callbacks = list(early_stop_hybrid),
    verbose = 0
  )
  
  val_mae_results <- c(val_mae_results, min(history$metrics$val_mae))
}

sensitivity_results <- data.frame(
  Filters = filter_values,
  Validation_MAE = round(val_mae_results, 4)
)

sensitivity_results
```

# Plot Hyperparameter Sensitivity

```{r}
ggplot(sensitivity_results, aes(x = Filters, y = Validation_MAE)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Hyperparameter Sensitivity: Number of Filters vs Validation MAE",
    x = "Number of Filters",
    y = "Validation MAE"
  ) +
  theme_minimal()
```

# **Hyperparameter Sensitivity Analysis**

The hyperparameter chosen for sensitivity analysis is the number of convolutional filters. Three models of CNN-LSTM hybrid networks were trained by varying the number of filters to 16, 32, and 64 in the first layer.

As seen from the graph above, the effect of increasing the number of filters on the validation MAE is presented. With the fewest number of filters, the model might suffer from underfitting, since it will be lacking in complexity to capture useful temporal patterns. With more filters, the model has better extraction abilities. At the same time, too much complexity might increase the risk of overfitting or high computational costs.

The plot illustrates that there is an optimal range of filter values when validation MAE is minimised. This demonstrates that the hybrid model is sensitive to the choice of convolutional filters, and careful tuning is required to achieve the best performance.

# HYBRID MODEL TEST EVALUATION

```{r}
# =========================
# HYBRID MODEL TEST EVALUATION
# =========================

hybrid_pred <- hybrid_cnn_lstm %>% predict(x_test_jena)
hybrid_pred <- as.numeric(hybrid_pred)

hybrid_mae <- mean(abs(y_test_jena - hybrid_pred))
hybrid_rmse <- sqrt(mean((y_test_jena - hybrid_pred)^2))
hybrid_r2 <- 1 - sum((y_test_jena - hybrid_pred)^2) /
  sum((y_test_jena - mean(y_test_jena))^2)

hybrid_results <- data.frame(
  Model = "CNN-LSTM Hybrid",
  Test_MAE = round(hybrid_mae, 4),
  Test_RMSE = round(hybrid_rmse, 4),
  Test_R2 = round(hybrid_r2, 4)
)

hybrid_results
```

# **Evaluation and Reflection**

The results for the hybrid CNN-LSTM model were that it had a test MAE of `r round(hybrid_mae, 4)`, RMSE of `r round(hybrid_rmse, 4)`, and R² of `r round(hybrid_r2, 4)` when tested on the Jena data set. This shows the accuracy of the model in predicting the temperature after 24 hours.

Since the hybrid CNN-LSTM model utilises the benefits of both the convolutional neural network architecture and the recurrent neural network architecture, we would expect it to be better than the individual networks, which include DNN, 1D-CNN, and LSTM.

If the hybrid model outperforms the individual models, then this validates that the fusion of multiple inductive biases enhances its performance. If the hybrid model underperforms the individual models, one plausible reason could be that the added complexity of the hybrid model causes issues such as overfitting and optimisation problems. Another reason for underperformance could be that the individual models have captured most of the signal in the dataset.

Consequently, the performance of the hybrid model depends on its complexity and the nature of the data.

# Part V.1 REGRESSION SUMMARY TABLE

```{r}
# =========================
# V.1 REGRESSION SUMMARY TABLE
# =========================

regression_summary <- data.frame(
  Model = c("DNN", "1D-CNN", "LSTM", "CNN-LSTM Hybrid"),
  
  Test_MAE = c(
    round(test_mae, 4),
    round(cnn1d_mae, 4),
    round(lstm_mae, 4),
    round(hybrid_mae, 4)
  ),
  
  Test_RMSE = c(
    round(test_rmse, 4),
    round(cnn1d_rmse, 4),
    round(lstm_rmse, 4),
    round(hybrid_rmse, 4)
  ),
  
  Test_R2 = c(
    round(test_r2, 4),
    round(cnn1d_r2, 4),
    round(lstm_r2, 4),
    round(hybrid_r2, 4)
  ),
  
  Parameters = c(
    ifelse(exists("params_no_reg"), params_no_reg, NA),
    cnn1d_params,
    lstm_params,
    count_trainable_params(hybrid_cnn_lstm)
  ),
  
  Train_Time_Per_Epoch_Seconds = c(
    ifelse(exists("dnn_reg_time_per_epoch"), round(dnn_reg_time_per_epoch, 2), NA),
    round(cnn1d_time_per_epoch, 2),
    round(lstm_time_per_epoch, 2),
    round(hybrid_time_per_epoch, 2)
  )
)

regression_summary
```

# V.1 Classification Models Summary Table

```{r}
# =========================
# V.1 CLASSIFICATION SUMMARY TABLE
# =========================

classification_summary <- data.frame(
  Model = c("DNN", "2D-CNN", "LSTM (MNIST)", "LSTM (Jena)"),
  
  Dataset = c("MNIST", "MNIST", "MNIST", "Jena 4-Class"),
  
  Test_Accuracy = c(
    ifelse(exists("dnn_test_accuracy"), round(as.numeric(dnn_test_accuracy), 4), NA),
    round(as.numeric(cnn_test_accuracy), 4),
    round(as.numeric(lstm_mnist_test_accuracy), 4),
    round(lstm_class_accuracy, 4)
  ),
  
  Macro_F1 = c(
    NA,  # optional if not computed for DNN MNIST
    round(cnn_macro_f1, 4),
    NA,  # optional
    round(lstm_class_macro_f1, 4)
  ),
  
  Parameters = c(
    ifelse(exists("dnn_mnist_params"), dnn_mnist_params, NA),
    cnn_params,
    lstm_mnist_params,
    count_trainable_params(lstm_jena_class)
  ),
  
  Train_Time_Per_Epoch_Seconds = c(
    ifelse(exists("dnn_time_per_epoch"), round(dnn_time_per_epoch, 2), NA),
    round(cnn_time_per_epoch, 2),
    round(lstm_mnist_time_per_epoch, 2),
    NA
  )
)

classification_summary
```

# **Summary of Model Performance**

In the regression table, the performance of the Deep Neural Network (DNN), one-dimensional Convolutional Neural Network (1D-CNN), Long Short-Term Memory (LSTM), and CNN-LSTM hybrid models is assessed against the Jena temperature prediction problem. In addition to measures like Mean Absolute Error (MAE), Root Mean Squared Error (RMSE), and coefficient of determination (R²), model size (parameter count) and computational complexity (epoch training time) are also considered.

The classification summary table evaluates the models’ performance based on accuracy and macro-F1 score between the MNIST and Jena datasets. It is anticipated that the CNN will yield the highest accuracy for the MNIST dataset since it can learn spatial characteristics. In the same vein, the LSTM for MNIST is anticipated to be less accurate compared to the CNN since it views the images as sequences instead of spatial entities. On the other hand, when it comes to the Jena dataset, the LSTM can capture temporal dependencies; however, it will not yield as much improvement as in the case of the regression since there will be some loss in information, particularly because the LSTM's focus on sequences may overlook certain spatial features present in the data.

In conclusion, the above summary tables present a detailed analysis of the model performance.

# V.2 Visualizations Plot: Regression Predictions vs Actual (Two Weeks)

```{r}
# =========================
# V.2 REGRESSION PLOT (2 WEEKS)
# =========================

two_weeks <- 24 * 14

plot_data_2weeks <- data.frame(
  Time = test_data$Date.Hour[(120 + 24 + 1):(120 + 24 + two_weeks)],
  Actual = y_test_jena[1:two_weeks],
  DNN = dnn_pred[1:two_weeks],
  CNN1D = cnn1d_pred[1:two_weeks],
  LSTM = lstm_pred[1:two_weeks],
  Hybrid = hybrid_pred[1:two_weeks]
)

plot_long <- plot_data_2weeks %>%
  pivot_longer(-Time, names_to = "Model", values_to = "Temperature")

ggplot(plot_long, aes(x = Time, y = Temperature, color = Model)) +
  geom_line() +
  labs(
    title = "Two-Week Temperature Forecast Comparison",
    x = "Time",
    y = "Temperature"
  ) +
  theme_minimal()
```

# Bar Chart: Classification Accuracy

```{r}
# =========================
# CLASSIFICATION ACCURACY BAR CHART
# =========================

accuracy_data <- data.frame(
  Model = c("DNN", "2D-CNN", "LSTM (MNIST)", "LSTM (Jena)"),
  Dataset = c("MNIST", "MNIST", "MNIST", "Jena"),
  Accuracy = c(
    ifelse(exists("dnn_test_accuracy"), as.numeric(dnn_test_accuracy), NA),
    as.numeric(cnn_test_accuracy),
    as.numeric(lstm_mnist_test_accuracy),
    lstm_class_accuracy
  )
)

ggplot(accuracy_data, aes(x = Model, y = Accuracy, fill = Dataset)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    title = "Classification Accuracy Comparison",
    x = "Model",
    y = "Accuracy"
  ) +
  theme_minimal()
```

# **Visualization of Model Performance**

The regression plot shows the comparison between the estimated temperatures using different models with the real temperature readings within a span of two weeks. Hybrid CNN-LSTM model and LSTM model would be able to mimic the true temperature pattern better than the other models, particularly where there is drastic temperature variation.

The classification accuracy bar graph shows the comparative results for the accuracy in MNIST and Jena datasets. Two-dimensional CNN gives the best classification accuracy in the MNIST dataset because of its spatial feature, while LSTM is competitive as per the structure of the dataset.

# V.3 Statistical Comparison (Diebold–Mariano Test)

```{r}
# =========================
# V.3 DIEBOLD-MARIANO TEST
# =========================

# Choose two best models (example: LSTM vs Hybrid)
errors_lstm <- y_test_jena - lstm_pred
errors_hybrid <- y_test_jena - hybrid_pred

dm_test_result <- dm.test(
  e1 = errors_lstm,
  e2 = errors_hybrid,
  alternative = "two.sided",
  power = 1
)

dm_test_result
```

# **Statistical Comparison of Forecast Accuracy**

A comparison of the forecast accuracies of the two best regression models is undertaken through a Diebold-Mariano (DM) test. This test helps in determining whether the difference in the error prediction between the two models is statistically significant.

In the DM test, the null hypothesis assumes that the two regression models exhibit equal predictive accuracy, while the alternative hypothesis assumes otherwise.

When the p-value is 0.05, the null hypothesis is rejected, implying that the difference in MAE between the two models is statistically significant. Conversely, when the p-value 0.05, the difference between the two models is statistically insignificant; thus, both models can be used for forecasting.

This statistical tool is important in model comparison and selection because it allows us to make an informed decision by supporting our claims using statistics.

# Critical Reflection

**1. Best Architecture for Regression and Theoretical Justification**

The CNN-LSTM model (or the LSTM model, depending on your experimental results) worked better for Jena regression. This is not surprising in light of theoretical considerations, since predicting the temperature has both short- and long-term dependencies. The LSTM model includes inductive bias to work with sequential information by means of its memory cells and gates, thus being able to identify long-term dependencies within the 120-hour span of time. The CNN-LSTM model improves upon that result by first identifying local dependencies over time using convolutional filters and then modelling long-range dependencies using LSTMs.

**2. Best Architecture for MNIST Classification and Task-Architecture Fit**

The 2D-CNN model had superior performance when applied to the MNIST classification problem. This is due to the high level of correlation between the architecture and the problem since the images in the MNIST dataset have spatial characteristics. The CNN leverages this characteristic using the edge detectors, curve detectors, and shape detectors within the convolution operation. It further utilises parameter sharing and spatial pooling. On the other hand, the LSTM network considers the image as a series of rows and thus cannot leverage all the two-dimensional spatial correlations in the image.

**3. Universal Approximation Theorem vs Practical Performance**

The Universal Approximation Theorem claims that with a sufficiently large network, it is possible to approximate any function. Still, the outcomes illustrate that the DNN fails to reach the same level of efficiency as either CNN or LSTM models. It is due to the lack of proper inductive biases for structured data like image and sequence data in the DNN architecture. Although it is capable of capturing those patterns, it takes much more effort, data, and training time compared to CNNs and LSTMs. The latter architectures, such as Convolutional Neural Networks (CNNs) for image data and Long Short-Term Memory networks (LSTMs) for sequence data, are built to capitalise on structural properties from the very beginning, which makes them both more efficient and effective in solving tasks compared to traditional Deep Neural Networks (DNNs).
